---
title: "Final Project"
author: "Daniel Justice"
date: "3/20/2020"
output:
  html_document:
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Solar system small body analysis

The data presented here is a sample collected from NASA's JPL Small-Body Database.
They represent smaller objects in the Solar System such as comets, asteriods, and moons.
I chose this data set because I am curious about space science, and this is an opportunity to learn more about the subject while fulfilling the requirements of the assignment.
Please forgive any inaccurate statements about orbital mechanics!
There are **22,458 cases** in my sample.

### Variables

|Title|Description|Type|
|---|---|---|
|full_name|Full Name|string|
|a| [au] semi-major axis|float64|
|e| eccentricity|float64|
|i| [deg] inclination|float64|
|w| [deg] perihelion degree|float64|
|om| [deg] node degree (longitude of the ascending node)|float64|
|ma| [deg] M mean anomaly degree|float64|
|q| [au] perihelion distance|float64|
|ad| [au] Q aphelion distance|float64|
|per_y| period years|float64|
|H| absolute magnitude|float64|
|moid| [au] Earth MOID (minimum orbit intersection distance)|float64|
|orbit_id|Orbit solution ID|string|
|class|Orbit class|string|

All of the `float64` data types are **quantitative** variables, and the `string` data types `orbit_id` and `class` are **categorical**.

### Oribit Classes

Below is a table of orbit classes.  There is a wealth of information on the [CNEOS website](https://cneos.jpl.nasa.gov/) that provides much more detail than I can here.

|class|name|num|
|---|:---|---|
|AMO|Amor|8267|
|APO|Apollo|12308|
|ATE|Aten|1684|
|ETc|Encke-type Comet|1|
|HTC|Halley-type Comet|30|
|IEO|Atira|21|
|JFC|Jupiter-family Comet*|10|
|JFc|Jupiter-family Comet|137|

JFC -> Jupiter-family comet, classical definition (P < 20 y).  
JFc -> Jupiter-family comet, as defined by Levison and Duncan (2 < TJupiter < 3).

```{r, echo=FALSE, results=FALSE}
mydata = read.csv("../neo.csv")
summary(mydata)
```
    
## Eccentricity

The first quantitative variable I would like to explore is labeled `e` for eccecentricity.
This variable takes on values between 0 and 1 where 0 is a perfectly circular orbit and close to 1 flattens the orbit into line.
The distribution appears to be close to **normal** and slighly **skewed to the right**.

```{r, echo=FALSE}
e_mean = round(mean(mydata$e), digits = 3)
e_median = round(median(mydata$e), digits = 3)
```

The **mean** is `r e_mean`.  
The **median** is `r e_median`.

```{r, echo=FALSE}
hist(mydata$e, main="Histogram of Eccentricity", xlab = "eccentricity")
```

```{r, echo=FALSE, results=FALSE}
summary(mydata$e)
```

```{r, echo=FALSE}
# Outliers Q3 + 1.5 * IQR > value < Q1 - 1.5 * IQR
q1 = quantile(mydata$e, 0.25)
q3 = quantile(mydata$e, 0.75)
iqr = q3 - q1
low_outliers <- mydata[which(mydata$e < q1 - (1.5 * iqr)),]
high_outliers <- mydata[which(mydata$e > q3 + (1.5 * iqr)),]
```


The **IQR** is `r round(iqr, digits=3)`.  
The **standard deviation** $s$ is `r round(sd(mydata$a), digits=3)`.  

There are `r length(low_outliers$e)` **outlier** values les than `r round(q1 - (1.5 * iqr), digits=3)`  
There are `r length(high_outliers$e)` **outlier** values greater than `r round(q3 + (1.5 * iqr), digits=3)`  


```{r, echo=FALSE}
boxplot(mydata$e, main="Eccentricity Boxplot")
```

## Semi-major axis

The other quantitative variable I would like to explore is the semi-major axis.
This is the larger of the two axis of an ellipse (for `e` > 0).
This distribution has a log-normal shape and is **skewed to the right** with a much longer tail than the previous graph.

The **mean** is `r e_mean`.  
The **median** is `r e_median`.

Please note the *logarithmic scale* of the y-axis in the following graph.
```{r, echo=FALSE}
hist(log(mydata$a), main="Histogram of Semi-Major Axis", xlab = "semi-major axis (AU)")
```

```{r, echo=FALSE, results=FALSE}
summary(mydata$a)
```

```{r, echo=FALSE}
# Outliers Q3 + 1.5 * IQR > value < Q1 - 1.5 * IQR
q1 = quantile(mydata$a, 0.25)
q3 = quantile(mydata$a, 0.75)
iqr = q3 - q1
low_outliers <- mydata[which(mydata$a < q1 - (1.5 * iqr)),]
high_outliers <- mydata[which(mydata$a > q3 + (1.5 * iqr)),]
```

The **IQR** is `r round(iqr, digits=3)`.  
The **standard deviation** $s$ is `r round(sd(mydata$a), digits=3)`.  

There are `r length(low_outliers$e)` **outlier** values les than `r round(q1 - (1.5 * iqr), digits=3)`  
There are `r length(high_outliers$e)` **outlier** values greater than `r round(q3 + (1.5 * iqr), digits=3)`  

```{r, echo=FALSE}
boxplot(mydata$a, main="Eccentricity Boxplot", log="y", xlab="semi-major axis (AU)")
```
```{r, echo=FALSE}
library(hexbin)
# plot(mydata$e, mydata$a, log="y", ylab="semi-major axis", xlab="eccentricity")

hbin <- hexbin(mydata$e, mydata$a, xbins=50)
plot(hbin, main="Semi-major axis vs Eccentricity", ylab="semi-major axis (AU)\n", xlab="eccentricity")
```


## Resources

- https://ssd.jpl.nasa.gov/sbdb_query.cgi 
- https://earthobservatory.nasa.gov/features/OrbitsCatalog
- https://www.airuniversity.af.edu/Portals/10/AUPress/Books/AU-18.PDF
- https://cneos.jpl.nasa.gov/about/neo_groups.html
- https://cneos.jpl.nasa.gov/glossary/
- https://www.statmethods.net/graphs/scatterplot.html